---
title: "Computational musicology 2025"
author: "Mees"
date: "2025-02-21"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)
library(Cairo)
library(ggtext)
```

### Introduction of my storyboard

Welcome to my storyboard for the computational musicology course. With this course 1 made 2 songs. I created these songs using the generative AI tool Stable Audio. I simply wrote a short prompt for each track to guide the style and mood.

**Track 1**  
Prompt:  
```
3 minutes Pixelwave, jazz fusion, retro 8-bit melodies, smooth synths, glitch effects, complex rhythms, futuristic vibes, melodic improvisation, digital textures, ambient soundscapes, quirky transitions
```

```{r, results='asis', echo=FALSE}
cat('<audio controls>
  <source src="mees-k-1.mp3" type="audio/mpeg">
</audio>')
```

**Track 2**  
Prompt:  
```
Classical Grunge, raw grunge guitar, symphonic strings, orchestral arrangements, distorted power chords, melancholic melodies, classical structure, gritty atmosphere, dynamic contrast, emotional intensity.
```

```{r, results='asis', echo=FALSE}
cat('<audio controls>
  <source src="mees-k-2.mp3" type="audio/mpeg">
</audio>')
```

### Visualisation of the corpus features

```{r pressure, echo=FALSE}
df <- read.csv("compmus2025.csv", stringsAsFactors = FALSE)

df_long <- df %>%
  pivot_longer(
    cols = -filename,
    names_to = "feature",
    values_to = "value"
  ) %>%
  group_by(feature) %>%
  mutate(norm_value = (value - min(value)) / (max(value) - min(value))) %>%
  ungroup()

pl <- ggplot() +
  geom_tile(data = df_long, aes(x = feature, y = filename, fill = norm_value,
                                text = paste("Feature:", feature,
                                             "<br>Song:", filename,
                                             "<br>Original value:", value)),
            color = "grey90") +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(x = "Feature", y = "Song") +
  scale_y_discrete(labels = function(x) {
    ifelse(x %in% c("mees-k-1", "mees-k-2"),
           paste0("<span style='color:red;'>", x, "</span>"),
           x)
  }) +
  theme_minimal() +
  theme(
    axis.text.y = element_markdown(size = 7),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplotly(pl, tooltip = "text")
```

*** 

This heatmap shows the normalized values of different features for 90 songs. Each tile’s color represents the scaled value (from 0 to 1) of a feature, making comparisons fair even when the original numbers differ a lot. Hover over a tile to see the real value. My tracks, "mees-k-1" and "mees-k-2", are highlighted in red.


### Conclusion
In conclusion, there is significant variance among the tracks produced by my peers. I look forward to sharing more detailed visualizations and insights as I continue exploring these musical features.
