---
title: "Computational musicology 2025"
author: "Mees"
date: "2025-03-30"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)
library(Cairo)
library(ggtext)
source("compmus.R")
library(rjson)
library(tidymodels)
library(ggdendro)
```

### Introduction of my storyboard

Welcome to my storyboard for the computational musicology course. With this course I made 2 songs. I created these songs using the generative AI tool Stable Audio. I simply wrote a short prompt for each track to guide the style and mood.

**Track 1**  
Prompt:  
```
3 minutes Pixelwave, jazz fusion, retro 8-bit melodies, smooth synths, glitch effects, complex rhythms, futuristic vibes, melodic improvisation, digital textures, ambient soundscapes, quirky transitions
```

```{r, results='asis', echo=FALSE}
cat('<audio controls>
  <source src="mees-k-1.mp3" type="audio/mpeg">
</audio>')
```

**Track 2**  
Prompt:  
```
Classical Grunge, raw grunge guitar, symphonic strings, orchestral arrangements, distorted power chords, melancholic melodies, classical structure, gritty atmosphere, dynamic contrast, emotional intensity.
```

```{r, results='asis', echo=FALSE}
cat('<audio controls>
  <source src="mees-k-2.mp3" type="audio/mpeg">
</audio>')
```

I wanted to create some unique songs that would maybe show up as outliers in the class corpus. I've tried to mix a lot of genres and chaotic terms into my prompts. 

### Visualisation of the corpus features

```{r pressure, echo=FALSE}
df <- read.csv("compmus2025.csv", stringsAsFactors = FALSE)

df_long <- df %>%
  pivot_longer(
    cols = -filename,
    names_to = "feature",
    values_to = "value"
  ) %>%
  group_by(feature) %>%
  mutate(norm_value = (value - min(value)) / (max(value) - min(value))) %>%
  ungroup()

pl <- ggplot() +
  geom_tile(data = df_long, aes(x = feature, y = filename, fill = norm_value,
                                text = paste("Feature:", feature,
                                             "<br>Song:", filename,
                                             "<br>Original value:", value)),
            color = "grey90") +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(x = "Feature", y = "Song") +
  scale_y_discrete(labels = function(x) {
    ifelse(x %in% c("mees-k-1", "mees-k-2"),
           paste0("<span style='color:red;'>", x, "</span>"),
           x)
  }) +
  theme_minimal() +
  theme(
    axis.text.y = element_markdown(size = 7),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplotly(pl, tooltip = "text")
```

***
This heatmap shows the normalized values of different features for 90 songs. Each tile’s color represents the scaled value (from 0 to 1) of a feature, making comparisons fair even when the original numbers differ a lot. Hover over a tile to see the real value. My tracks, "mees-k-1" and "mees-k-2", are highlighted in red. My 2 songs don't really have extreme values compared to the other songs.

### Chromagram of my 2 songs

```{r chromagram, echo=FALSE}
chroma1 <- compmus_chroma("mees-k-1.json", norm = "identity") %>% 
  mutate(song = "mees-k-1")
chroma2 <- compmus_chroma("mees-k-2.json", norm = "identity") %>% 
  mutate(song = "mees-k-2")

chromagram <- bind_rows(chroma1, chroma2)

pl <- ggplot(chromagram, aes(x = time, y = factor(pc), fill = value)) +
  geom_tile() +
  facet_wrap(~ song, ncol = 1) +
  scale_fill_viridis_c() +
  labs(x = "Time (s)", y = "Pitch Class", title = "Chromagram of My Songs") +
  theme_minimal()

ggplotly(pl)
```

***
These are the chromagrams of my two songs. You can see that my first song has a lot of variance in the pitch classes. If you listen to "mees-k-1," you can hear where this comes from. With my second song, a more defined pattern can be seen. If you listen to that song, it may also sound less random and more structured.

### Self-similarity chroma matrix of my first song

```{r selfsim, echo=FALSE}
chroma_df <- compmus_chroma("mees-k-1.json", norm = "identity")
sim_chroma <- compmus_self_similarity(chroma_df, value, distance = "euclidean")

selfsim <- ggplot(sim_chroma, aes(x = xtime, y = ytime, fill = d)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "Chroma-based Self-Similarity for mees-k-1", x = "Time (s)", y = "Time (s)", fill = NULL) +
  theme_minimal()

selfsim
```

***

To get a closer look at the randomness of my first song, we can examine the self-similarity matrix. We observe that there are no continuous diagonal lines running through the plot, which indicates that there are no obvious repeating patterns in the song's chroma. However, you can see a slight checkerboard pattern. This means that even though the song doesn't have clear, repeating sections, some parts of it sound very similar to themselves.

### Self-similarity timbre matrix of my second song

```{r selfsimtimbre, echo=FALSE}
"mees-k-2.json" |>
  compmus_mfccs(norm = "identity") |>
  compmus_self_similarity(
    feature = mfcc,
    distance = "euclidean"
  ) |>
  ggplot(aes(x = xtime, y = ytime, fill = d)) +
  geom_raster() +
  scale_fill_viridis_c() +
  geom_tile() +
  labs(x = "Time (s)", y = "Time (s)", fill = NULL, 
       title = "Timbre-based Self-Similarity Matrix for mees-k-2") +
  theme_minimal()
```

***

By observing the self-similarity matrix for timbre, we can see that the song contains two large patches with high similarity. This indicates that these parts of the song have many similar timbre-based features. When listening to the song, we can also hear that these segments prominently feature the electric guitar.


### Key and Chord estimation of my 2 songs

```{r keyesti, echo=FALSE}
#From chordogram.R
chordogram <- readRDS("chordogram.rds")
ggplotly(chordogram)
```

***

These chordograms are very similar to the chromagrams we saw earlier. In this plot, you can see that my first song uses a wide variety of chords, with the final part being especially chaotic. In my second song, the structure is clearer, with a few chords repeated throughout. The chordogram of my second song also shows the sections where the electric guitar is present.


### Histogram of tempo of the class corpus

```{r tempohist, echo=FALSE}
class_corpus <- read_csv("compmus2025.csv")
tempo_histogram <- ggplot(class_corpus, aes(x = tempo)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black") +
  labs(
    title = "Histogram of Tempo for the Class Corpus",
    x = "Tempo (BPM)",
    y = "Frequency"
  ) +
  theme_minimal()

ggplotly(tempo_histogram)
```

***
This plot shows a histogram of the tempos in the class corpus. My songs have tempos of 83 and 110 bpm, which are similar to those of the other songs. There are also five songs with very low tempos, estimated to be around 30 bpm. Because this value is rare, I believe the software may have made a mistake in estimating these tempos.

### Tempograms of my own 2 songs

```{r tempogram, echo=FALSE}
#From tempogram.R
tempogram <- readRDS("tempogram.rds")
ggplotly(tempogram)
#83 110
```

***
These are the tempograms of my own 2 songs. These plots are really not that interesting because there are no tempo changes in my songs. 


### Hierarchical clustering of the class corpus

```{r dendro, echo=FALSE}

get_conf_mat <- function(fit) {
  outcome <- .get_tune_outcome_names(fit)
  fit |> 
    collect_predictions() |> 
    conf_mat(truth = outcome, estimate = .pred_class)
}  

get_pr <- function(fit) {
  fit |> 
    conf_mat_resampled() |> 
    group_by(Prediction) |> mutate(precision = Freq / sum(Freq)) |> 
    group_by(Truth) |> mutate(recall = Freq / sum(Freq)) |> 
    ungroup() |> filter(Prediction == Truth) |> 
    select(class = Prediction, precision, recall)
}  

compmus2025 <- read_csv("compmus2025.csv")

cluster_juice <-
  recipe(
    filename ~
      arousal +
      danceability +
      instrumentalness +
      tempo +
      valence,
    data = compmus2025
  ) |>
  step_center(all_predictors()) |>
  step_scale(all_predictors()) |> 
  prep(compmus2025) |>
  juice() |>
  column_to_rownames("filename")


compmus_dist <- dist(cluster_juice, method = "euclidean")


hc <- hclust(compmus_dist, method = "single")
ddata <- dendro_data(hc)

my_songs <- c("mees-k-1", "mees-k-2")

ddata$labels <- ddata$labels %>% 
  mutate(label = if_else(label %in% my_songs,
                         paste0("<span style='color:red;'>", label, "</span>"),
                         label))

dendro <- ggplot() +
  geom_segment(data = ddata$segments, 
               aes(x = x, y = y, xend = xend, yend = yend)) +
  ggtext::geom_richtext(data = ddata$labels, 
                        aes(x = x, y = y, label = label),
                        hjust = 1, angle = 90, 
                        fill = NA, label.color = NA, size = 3) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(plot.margin = margin(t = 10, r = 50, b = 50, l = 10))


dendro
```

***

My first goal was to create outliers in the class corpus, making the clustering more interesting. My songs are indicated by the red text. However, in this dendrogram, we can see that my songs are very similar to other songs in the corpus, so my goal was not achieved. One outlier in the set is the song "lo-l-1." You can listen to it below.

```{r, results='asis', echo=FALSE}
cat('<audio controls>
  <source src="lo-l-1.mp3" type="audio/mpeg">
</audio>')
```

If you listen to the song, you'll notice that it is a unique track with a slow pace and many influences from classical music.

### Conclusion

Overall, my analyses show that while my songs have some unique features, they are not extreme outliers in the class corpus. The chromagrams and self-similarity matrices reveal differences between my two tracks, one being more varied and the other more structured, but the clustering results indicate that my songs still group with many other similar tracks. If I had the chance to try this again I think I need to focus more on the high level features of my songs to stand out better in the class corpus.

I really enjoyed exploring and listening to the songs created by my classmates. It was fun to see so many creative approaches and learn from their unique styles.